---
title: "EcoCluster: Bayesian Clustering using Truncated Stick-Breaking priors "
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EcoCluster}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette illustrates three different types of Bayesian clustering methods using truncated stick-breaking priors: mixture models, stochastic block models (SBM), and species archetype models (SAM). We apply these three clustering methods to simulated data to showcase how the number of groups can be successfully recovered.

# Mixture model

We start by generating simulated data:
```{r,eval=TRUE}
rm(list=ls(all=TRUE))
set.seed(1)

#general settings
nloc=300 #number of locations
nspp=50  #number of species
ngroup=5 #number of groups

#set parameters
z=sample(1:ngroup,size=nloc,replace=T)         #cluster assignment of each location
phi=matrix(rbeta(ngroup*nspp,1,1),ngroup,nspp) #species composition of each cluster  

#generate data
y=matrix(NA,nloc,nspp)
for (i in 1:nloc){
  phi1=phi[z[i],]
  y[i,]=rbinom(nspp,size=1,prob=phi1)
}
colnames(y)=paste0('spp',1:nspp)
rownames(y)=paste0('loc',1:nloc)

head(y[,1:13])
```

To obtain posterior samples from the mixture model parameters, we rely on the function `mixture.gibbs.main.func`. The user needs to specify the following arguments:

* `dat`:    binary matrix containing the data (rows are locations and columns are species) 
* `ngroup`: maximum number of groups
* `ngibbs`: number of iterations for the Gibbs sampler
* `burnin`: number of iterations to be discarded as burn-in.

```{r,eval=TRUE,results='hide'}
library('EcoCluster')
```